Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-07-26 manuscript/manuscript.qmd
included Bob's edits and comments

Real-world evidence (RWE) is increasingly used to complement findings from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials for non-oncology indications. Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology. ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the interpretation of results from emulated oncology trials using RWD.

@@

However, the comparability and transportability of results derived between non-interventional studies and RCTs depends on multiple factors. Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of non-interventional studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when non-interventional studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, non-interventional studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.

@@

The focus of ENCORE is to efficiently explore in what contexts non-interventional studies can or cannot yield similar results compared to oncology RCTs. Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years. After review and collaboration with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).

@@

-   Feasibility and clinical relevance (latter was defined as treatment or paradigm-changing trials or such that challenged existing treatment policies)

The rationale and operationalization for each criterion is listed in detail in @tbl-criteria. We mainly considered pivotal large late-phase RCTs that were initiated after 2011 because treatment guidelines for included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of newly-approved therapies in routine care, patients are less likely to receive outdated treatment regimens in clinical practice. Conversely, we excluded trials with results published too recently to allow for substantive medical product uptake and follow-up time accrual in real-world databases used for this project. We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a clinically relevant basis.

@@

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts. The corresponding selection process is illustrated in the CONSORT diagrams (**Supplementary Figures 1-4**) and runner up candidate trials are listed in **Supplementary Table 1**. The majority of shortlisted trials represent indications for advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer patients because a large proportion of drug development efforts have recently focused on these settings. A key objective that we aim to explore with the selected trials is to evaluate how stage (early, late), line of therapy (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments after applying all relevant eligibility criteria suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.

@@

The ENCORE project will use data from four U.S.-based oncology-specific EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. A detailed description and sampling methodology will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication given the specificity of data elements required and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in randomly assigned order). If more than one database is considered fit-for-use for a respective trial emulation, the most suitable analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

The distribution of patient characteristics, by exposure status, will be examined before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline, and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design or analytic choices based on known outcome information.

@@

Causal analyses of non-interventional data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results. Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival time differences) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively. The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart. For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.

@@

While the MAR assumption is a strong assumption to hold across all considered covariates, it has been shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), generally only mechanisms in which a covariate causes its own missingness lead to significant bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain otherwise eligible patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses, although such tradeoffs must be carefully evaluated within the clinical and study contexts.

>>>>

Real-world evidence (RWE) is increasingly used to complement findings from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice. Advances in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose. The RCT DUPLICATE initiative has previously evaluated when RWE can most appropriately draw causal conclusions by emulating trials for non-oncology indications. Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology. ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. It will place special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis. Pre-specified criteria will assess agreement of treatment effect estimates between RCTs and their respective emulations. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project aims to inform understanding of how measurement, design, and analytic decisions influence the interpretation of results from emulated oncology trials using RWD.

@@

However, the comparability and transportability of results derived between non-interventional studies and RCTs depends on multiple factors. Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized endpoint assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases would enhance increase confidence in the interpretability of non-interventional studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] evaluated when non-interventional studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that well-designed and conducted RCT findings reflect causal treatment effects. In settings where the RCT designs could be emulated well, non-interventional studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused on emulating trials in non-oncology settings using claims databases.

@@

The focus of ENCORE is to clarify when non-interventional studies can or cannot yield similar results compared to oncology RCTs. Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years. After review and collaboration with clinical and regulatory experts, four cancer types were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).

@@

-   Feasibility and clinical relevance (latter was defined as treatment or paradigm-changing trials or trials that challenged existing treatment policies)

The operationalization for each criterion is listed in detail in @tbl-criteria. We mainly considered pivotal large late-phase RCTs that were initiated after 2011 because treatment guidelines for included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of newly-approved therapies in routine care, patients are less likely to receive outdated treatment regimens in clinical practice. Conversely, we excluded trials with results published too recently to allow for substantive medical product uptake and follow-up time accrual in real-world databases used for this project. We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a clinically relevant basis.

@@

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts. The corresponding selection process is illustrated in the CONSORT diagrams (**Supplementary Figures 1-4**) and runner up candidate trials are listed in **Supplementary Table 1**. The majority of shortlisted trials represent indications for advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer patients because a large proportion of drug development efforts have recently focused on these settings. A key objective that we aim to explore with the selected trials is to evaluate how stage (early, late), line of therapy (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments after applying all relevant eligibility criteria (including formal power calculations and assessment of sufficient balance after propensity score matching or weighting) suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.

@@

The ENCORE project will use data from four U.S.-based oncology-specific EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. A detailed description and sampling methodology will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication given the specificity of data elements required and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in randomly assigned order). If more than one database is considered fit-for-use for a respective trial emulation, the most suitable analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using random effects (primary; weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) and fixed effects (secondary; weights reflect the inverse of the variance of each database study) models.[@nikolakopoulou2014interpret]

@@

The distribution of patient characteristics, by exposure status, will be examined before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced[@austin2009balance], exposure cohorts are conditionally exchangeable at baseline, and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design or analytic choices based on known outcome information.

@@

Causal analyses of non-interventional data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results. Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess whether the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$ with a 2-sided $\alpha$ of 0.05) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (log hazard ratios or median survival time differences) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively. The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart. For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.

@@

While MAR is a strong assumption to hold across all considered covariates, it has been shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), generally only mechanisms in which a covariate causes its own missingness lead to significant bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain otherwise eligible patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses, although such tradeoffs must be carefully evaluated within the clinical and study contexts.
